---
title: "Transmission model"
author: "Juan S Vargas"
date: "4/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MASS)
library(tidyverse)
```

## Prevalence of _A. caninum_ in domestic dogs

I have data on prevalence of Ancylostoma caninum in domestic dogs. We collected 110 samples in 2011/12. K. Gostic collected 92 samples, 57 from unknown individuals, 35 from 19 known individuals. In total she had 76 independent samples, 62 (82%) tested positive for _A. caninum_. 
J. Johnson had samples from 18 different dogs, 12 were positive for _A. caninum_. 
74 were positive for a strongylid, likely _A. caninum_. We can create a binomial glm to obtain confidence intervals on the estimated prevalence values combined. 

```{r dog prev}
prevdata <- read.csv('../data/prevalences.csv')
d <- prevdata[prevdata$host=='dog',]
m1 <- glm(cbind(pos_samp, neg_samp)~1, family = binomial, data = prevdata[prevdata$host=='dog',])
b <- c(coefficients(m1), confint(m1))
exp(b)/(1+exp(b))
```
The expected prevalence in dogs is 0.742 (95% CI: 0.669-0.806), including all areas. 

```{r ocelot prev}
m1 <- glm(prev~1, family = binomial, weights = n, data=prevdata[prevdata$host=='ocelot'&prevdata$location!='Esquinas',])
b <- c(coefficients(m1), confint(m1))
exp(b)/(1+exp(b))
```

The expected prevalence in ocelots is 27%, (7.6-56.4). For pumas it's 36.4% (18.6-57.2%).

```{r puma logistic model}
m1 <- glm(prev~1, family = binomial, weights = n, data=prevdata[prevdata$host=='puma',])
b <- c(coefficients(m1), confint(m1))
exp(b)/(1+exp(b))
```

These prevalences correspond to the presence of strongyle parasites. However, we did not find any evidence of *A. caninum* in wild feline samples. This can only be confirmed with molecular methods. Unfortunately, we only got good sequences from two ocelot and two puma samples, out of 8 positive puma samples, and 3 positive ocelot samples. We can use epidemiological models to gain a better understanding about the possibility of interspecific transmission.

## Multispecies host-parasite model

We create a model following the framework of Fenton et al. (2015). The model is a system with two species, who get infected with a parasite through contact with a shared pool of infectious parasite stages. This models allows estimating prevalence at equilibrium in both species, and to explore how this prevalence changes as a function of system parameters like the level of overlap and the density of hosts. The model is described by a system of ordinary differential equations as described in the text.

###Parameter estimates

To solve the model numerically, we need to estimate the frequency with which individual ocelots visit human-modified features (i.e. trails/roads).
We use the records of individual ocelots to determine the mean frequency of use of trails for each. 
```{r}
stn_data <- read_csv("../data/2019stationdata.csv") %>% filter(cam=="A") %>% 
  mutate(dt = lubridate::mdy(takedown_date))
read_csv("../data/ocelot_record_tbl.csv") %>% 
  mutate(Date=lubridate::mdy(Date)) %>% 
  filter(dog_station==T) %>% 
  group_by(Individual) %>%
  arrange(Date, .by_group=T) %>% 
  mutate(dtime=Date-lag(Date)) %>% 
  left_join(stn_data, by = join_by(Station==station)) %>% 
  add_count() %>% ungroup() %>% 
  mutate(dtime = if_else(n==1, dt-Date, dtime)) %>% 
  summarise(meandelta=weighted.mean(dtime, n,na.rm = T),
            mediandelta = median(dtime, na.rm=T)) 
```

We estimate that individual ocelots go on trails that dogs also use with a frequency of `r cat(1/4.61)` (0.024, SD:0.030)

```{r dog trail freq}
dogRecTbl <- read.csv("../data/dog_record_tbl")
dogRecTbl %>% group_by(Individual) %>% 
  left_join(stn_data, by = join_by(Station==station)) %>% 
  arrange(Station,Individual,Date) %>% 
  mutate(dtime = Date-lag(Date)) %>% 
  add_count() %>% 
  mutate(dtime = if_else(n==1,lubridate::mdy(takedown_date)-Date,dtime)) %>% 
  ungroup() %>% 
  summarise(meant = mean(dtime, na.rm=T), sd = sd(dtime, na.rm=T))
```
Similarly, the mean frequency for dogs is 0.088. 

```{r}
freqdata <- read_csv('ocelot_anthro_cap.csv') 
freqdata %>% 
  pivot_longer(cols = starts_with('Lp'), names_to='ocelot_id', names_ptypes=list(ocelot_id=factor())) %>% 
  filter(dog>0 ) %>% 
  group_by(ocelot_id) %>% summarise(pres=max(value>0)) %>% select(pres) %>% sum()
```
Of 22 ocelots identified, 18 or`r cat(round(18/22,3))` were seen at stations that also had presence of dogs or cattle.

```{r}
freqdata %>% filter(dog>0) %>% summarise(mean(dog/wk_days))
```
At stations where dogs pass, the frequency is high, 0.11/day, or 1 every 9 days. This is the mean, the median is 0.0372/day. If we assume a stable prevalence in dogs we can simplify the model and deal only with the visits to the common pool. The rate at which infectious feces appear in the environment available to ocelots would be the prevalence of hookworms in dogs, multiplied by the frequency of use of common trails. This is, $\lambda_1=0.742\times 0.111$, `r cat(format(0.742*0.111,digits=2,scientific=T), '/day',  sep='')`
We also need to estimate the rate at which infectious points disappear. For this, we use the survival rate of infectious larvae in the soil. 
Mean larval survival on grass in field conditions was 24 days (range 1-49) in the fall, 0 days in the winter (0-21), and 6.6 days over Spring/Summer (0-21) (Mark 1975). Mark (1975) also found that high temperature and rainfall favored survival. 
According to the literature, L3 larvae enter arrested development until they encounter a host, so they can survive long periods in the soil. They are, however, susceptible to dessication (although there is experimental evidence of survival through anhydriobiosis in Trichostrongyle nematodes; Lettini et al. 2006).
_In vitro_ larval development assays show 100% survival after 68 hours, or nearly three days (Colgrave et al. 2009). Longer term studies placed _A.caninum_ larvae in PBS medium and found a more or less linear reduction in the proportion of motile larvae able to migrate. These larvae were kept in the dark, at 22 C and survived for up to seven months (Kopp et al. 2008). There were reduction in migration ability of 50% after four months.
An experiment with _A. ceylanicum_ in hamsters estimated an establishment/survival rate of 24%, i.e. 1/4 of larvae uptaken develop to adulthood (larvae given orally) (Ribeiro Serafim et al. 2014).
_A. caninum_ larvae cultured in agar at 23 C. 88% were infective after 11 days, the rest were dead. Similar results at 30 C (McCoy). 
Komiya et al (1956) kept _A. caninum_ adults _in vitro_ alive for up to 6 weeks (males) and 12 weeks (female). The mean was 4 weeks for males and seven weeks for females, under physiological conditions. They also measured fecundity, estimated at a maximum of 7600 egg/female/day. This amount decreases progressively. 
Vicente et al. (2004) studied the epidemiology of _Ancylostoma_ in iberian lynx. 
_Ancylostoma_ has a prepatent period estimated at 13 to 27 days. Studies in seals estimate that healthy individuals are able to clear the infection after 50 days. This is slightly lower than the survival period of adult _A. caninum in vitro_ (mean was 4 months).
The survival of eggs was studied by Smith and Chad (1989). They determined survival and development of _A. duodenale_ and _N. americanus_ eggs at different temperatures. At 25 C, mortality rate would be 2.89E-3 /eggs/h, or 0.069/egg/d. In the same paper, the development rate was estimated as 

I take data from experimental studies to estimate infectivity--the probability of infection given contact--of _Ancylostoma spp._ in cats and dogs. I took the data from three sources, Scott (1928), Herrick (1928), and Okoshi & Murata (1967). The combined data suggests that the infectivity is roughly 97% for experimentally infected dogs. This value is likely lower in natural conditions. 
The infectivity of _A. caninum_ in cats was studied by Scott (1928) and Okoshi & Murata (1967). Okoshi & Murata report an infectivity between 5/14 for cutaneous infection and 4/13 for oral infection. For low doses, however, infectivity was lower than 1/6 cutaneously, and lower than 1/6 orally. 








